ECCCos from the Black Box: Letting Models speak for Themselves

### 2023-04-27 paper/paper.tex
attempting a run without normalising digits to between -1 and 1

The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}. For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state. That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}.

@@

\begin{minipage}[c]{0.40\textwidth}
  \centering
  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
  \captionof{figure}{Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.50\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, m, M$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $M$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},m)$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\section{Evaluation Framework}\label{conformity}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \caption{ECCCos from Black Boxes. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); }\label{fig:eccco}
\end{figure}

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 currently only one that works)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
\end{itemize}

>>>>

@@

\begin{minipage}[c]{0.40\textwidth}
  \centering
  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
  \captionof{figure}{Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.50\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\begin{minipage}[c]{\textwidth}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \captionof{figure}{Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
\end{minipage}

The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}. For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state. That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}. Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.

Finally, we search counterfactuals through gradient descent. Let $\mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$ denote our loss function defined in Equation~\ref{eq:eccco}. Then in each iteration, we first randomly draw $n_{\mathcal{B}}$ samples from the buffer $\mathcal{B}$ before updating the counterfactual state $\mathbf{z}^\prime$ by moving in the negative direction of that loss function. The search terminates once the convergence criterium is met or the maximum number of iterations $T$ has been exhausted. Note that the choice of convergence criterium has important implications on the final counterfactual. For more detail on this see Appendix~\ref{app:eccco}).

Figure~\ref{fig:eccco-mnist} presents ECCCos for the MNIST example from Section~\ref{background} for various Black Box models of increasing complexity from left to right: a simple Multi-Layer Perceptron (MLP); an Ensemble of MLPs, each of the same architecture as the single MLP; a Joint Energy Model (JEM) based on the same MLP architecture; and finally, an Ensemble of these JEMs. Since Deep Ensembles have an improved capacity for predictive uncertainty quantification and JEMs are explicitly trained to learn plausible representations of the input data, it is intuitive to see that the plausibility of counterfactuals visibly improves from left to right.

\section{Experiments}\label{conformity}

@@

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 currently only one that works)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
\end{itemize}
